---
title: "06 — Life-history data quality and parameter consistency"
subtitle: "Are biological inputs internally coherent, representative and transferable?"
author: "Reda M. Fahim"
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(ggplot2)
source("06_life_history_consistency.R")
```

## Why this comes before stock assessment

Life-history parameters are often treated as fixed inputs: a growth curve, a maturity ogive, a weight–length relationship and a value for natural mortality are assembled and passed to an assessment.

That can be dangerous. A parameter may be estimated correctly but still be inappropriate for the assessment if it represents a different stock, sex, period, sampling source, gear, ageing method or maturity protocol. The purpose of this module is therefore **not to estimate a definitive life-history parameter set**. It is to audit whether the biological evidence is internally coherent and sufficiently transferable for later use.

The simulated programme contains individual fish observations from 2016–2025 across month, area and commercial/survey sources. Deliberate inconsistencies are embedded so that the workflow must detect and interpret them rather than simply fit curves.

## 1. Start with record-level validity

```{r}
knitr::kable(issue_register, digits = 3, caption = "Life-history QC issue register")
```

The objective is not automatic deletion. Each flag points to a different action: unit verification, coding review, source-record checking, or sensitivity analysis.

## 2. Is biological sampling adequately supported?

Uneven sampling can make life-history differences appear biological when they are really differences in coverage.

```{r}
ggplot(support, aes(year, month, size = n_bio)) +
  geom_point(alpha = 0.65) +
  facet_grid(area ~ source) +
  scale_x_continuous(breaks = seq(2016, 2025, 3)) +
  labs(title = "Biological sampling support varies through space and time",
       subtitle = "Point size is the number of valid fish in each year × month × area × source stratum",
       x = NULL, y = "Month", size = "n") +
  theme_minimal(base_size = 12)
```

A parameter estimate should not be interpreted without knowing whether the underlying sampling frame is persistent, balanced and biologically representative.

## 3. Weight–length relationships: fit, then test stability

Weight–length relationships are commonly used to convert numbers-at-length to biomass or to fill missing weight observations. A high R² is not sufficient evidence that one relationship is valid across sexes, years or sampling sources.

```{r}
ggplot(valid[!valid$condition_outlier, ], aes(length_cm, weight_g)) +
  geom_point(alpha = 0.08, size = 0.6) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 3), se = FALSE, linewidth = 0.8) +
  facet_grid(sex ~ source) +
  labs(title = "Weight–length structure differs subtly by sex and source",
       subtitle = "Curves are descriptive; formal coefficients are estimated on the log scale",
       x = "Length (cm)", y = "Weight (g)") +
  theme_minimal(base_size = 12)
```

```{r}
knitr::kable(wl_summary, digits = 4,
  caption = "Sex- and source-specific log-linear weight–length relationships")
```

```{r}
ggplot(wl_year, aes(year, b, group = sex, linetype = sex)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  geom_hline(yintercept = 3, linetype = 3) +
  labs(title = "Temporal stability of the weight–length exponent",
       subtitle = "A stable overall relationship can still hide year-specific shifts",
       x = NULL, y = "Estimated exponent b", linetype = "Sex") +
  theme_minimal(base_size = 12)
```

## 4. Maturity: L50 is conditional on the observation process

The simulated data contain sex-specific maturity schedules and a deliberate recent commercial-sampling drift that under-represents small immature fish.

```{r}
maturity_plot_data <- valid
maturity_plot_data$length_bin <- floor(maturity_plot_data$length_cm / 2) * 2
mat_bin <- aggregate(mature ~ length_bin + sex + source, maturity_plot_data, mean)
mat_n <- aggregate(fish_id ~ length_bin + sex + source, maturity_plot_data, length)
names(mat_n)[4] <- "n"
mat_bin <- merge(mat_bin, mat_n, by = c("length_bin", "sex", "source"))
mat_bin <- mat_bin[mat_bin$n >= 20, ]

ggplot(mat_bin, aes(length_bin, mature, linetype = source)) +
  geom_line(linewidth = 0.9) +
  geom_point(aes(size = n), alpha = 0.65) +
  facet_wrap(~ sex) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Observed maturity ogives depend on sex and sampling source",
       subtitle = "Point size shows biological sample support within each 2-cm bin",
       x = "Length class (cm)", y = "Proportion mature", linetype = "Source", size = "n") +
  theme_minimal(base_size = 12)
```

```{r}
knitr::kable(maturity_summary, digits = 3, caption = "L50 estimates by sex and source")
```

```{r}
ggplot(l50_year, aes(year, L50, group = sex, linetype = sex)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2) +
  labs(title = "Annual L50 is a diagnostic time series, not just a single number",
       subtitle = "Changes may reflect biology, sampling composition or maturity-staging practice",
       x = NULL, y = "Estimated L50 (cm)", linetype = "Sex") +
  theme_minimal(base_size = 12)
```

```{r}
knitr::kable(late_l50, digits = 3, caption = "2024–2025 maturity sensitivity to sampling source")
```

A divergence between commercial and survey estimates is a reason to investigate the observation process before interpreting a temporal change in maturation.

## 5. Sex ratio is also design-sensitive

```{r}
ggplot(sex_ratio, aes(year, prop_female, group = area, linetype = area)) +
  geom_line(linewidth = 0.75) + geom_point(aes(size = n), alpha = 0.7) +
  facet_wrap(~ source) + geom_hline(yintercept = 0.5, linetype = 3) +
  labs(title = "Observed sex ratio varies with area and sampling source",
       subtitle = "Differences should not automatically be interpreted as population sex-ratio change",
       x = NULL, y = "Proportion female", linetype = "Area", size = "n") +
  theme_minimal(base_size = 12)
```

This matters whenever sex-specific growth, maturity or mortality inputs are used.

## 6. Spawning phenology: does the seasonal signal agree across components?

```{r}
ggplot(gsi_month, aes(month, gsi, group = sex, linetype = sex)) +
  geom_line(linewidth = 0.9) + geom_point(aes(size = n), alpha = 0.75) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Seasonal GSI identifies a coherent spawning window",
       subtitle = "Peak timing should be consistent with maturity staging and sampling coverage",
       x = "Month", y = "Mean GSI of mature fish", linetype = "Sex", size = "n") +
  theme_minimal(base_size = 12)
```

```{r}
knitr::kable(peak_month, digits = 2, caption = "Peak GSI month by sex")
```

```{r}
ggplot(gsi_area_month, aes(month, gsi, group = area, linetype = area)) +
  geom_line(linewidth = 0.8) + geom_point(size = 1.7) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Female spawning phenology by area", x = "Month", y = "Mean GSI", linetype = "Area") +
  theme_minimal(base_size = 12)
```

## 7. Growth parameters: estimation is not the same as compatibility

A simulated age subsample is used to fit sex-specific von Bertalanffy curves. The role of this step is to check coherence with the observed size range and with maturity information.

```{r}
knitr::kable(growth_summary, digits = 3,
  caption = "Sex-specific growth fits from the simulated ageing subset")
```

```{r}
ggplot(age_subset, aes(age_observed, length_cm)) +
  geom_jitter(width = 0.08, alpha = 0.07, size = 0.6) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.9) +
  facet_wrap(~ sex) +
  labs(title = "Age–length structure used to support growth estimates",
       subtitle = "The scatter and sample support matter as much as the fitted parameters",
       x = "Observed age", y = "Length (cm)") +
  theme_minimal(base_size = 12)
```

## 8. Parameter provenance audit

One of the most consequential pre-assessment tasks is deciding whether a candidate life-history parameter is actually compatible with the stock and data being assessed. The simulated catalogue intentionally mixes local/current, historical, adjacent-stock, sex-combined and default parameters.

```{r}
catalogue_display <- parameter_catalogue[, c("source_id", "parameter", "stock_unit", "sex",
  "period", "method", "value_1", "value_2", "units", "compatibility_score", "compatibility_class")]
knitr::kable(catalogue_display, digits = 3,
  caption = "Candidate life-history parameter provenance and compatibility")
```

```{r}
compat_counts <- as.data.frame(table(parameter_catalogue$parameter,
  parameter_catalogue$compatibility_class), stringsAsFactors = FALSE)
names(compat_counts) <- c("parameter", "compatibility", "n")
compat_counts <- compat_counts[compat_counts$n > 0, ]
ggplot(compat_counts, aes(compatibility, parameter, size = n)) +
  geom_point(alpha = 0.75) +
  labs(title = "Compatibility is part of the parameter, not an afterthought",
       subtitle = "A precise estimate from the wrong stock, period or sex can be less useful than a noisier local estimate",
       x = "Compatibility class", y = NULL, size = "Candidates") +
  theme_minimal(base_size = 12)
```

The compatibility score is deliberately simple and transparent. It is an **audit device**, not a universal scoring system. A real assessment should define stock-specific evidence rules and document expert judgement.

## 9. Natural mortality: compare estimators before choosing a sensitivity range

The previous version of this module used an illustrative low/central/high range. That was useful for demonstrating sensitivity but it did not answer an essential scientific question: **where should candidate values of M come from?**

This version calculates several commonly used empirical estimators explicitly. They are not interchangeable: each uses different biological information and embeds different assumptions.

```{r}
source("natural_mortality_methods.R")

# Illustrative mean annual habitat temperature for the simulated example.
# Replace this with stock-specific evidence when adapting the workflow.
temperature_C <- 17

tmax_by_sex <- aggregate(age_observed ~ sex, age_subset, max)
names(tmax_by_sex)[2] <- "tmax_observed"

M_inputs <- merge(growth_summary[, c("sex", "Linf", "K", "t0")], tmax_by_sex, by = "sex")
M_inputs$temperature_C <- temperature_C

M_scalar_comparison <- do.call(rbind, lapply(seq_len(nrow(M_inputs)), function(i) {
  d <- M_inputs[i, ]
  out <- compare_scalar_M(
    Linf_cm = d$Linf,
    K = d$K,
    temperature_C = d$temperature_C,
    tmax_years = d$tmax_observed
  )
  out$sex <- d$sex
  out
}))
M_scalar_comparison$annual_survival <- exp(-M_scalar_comparison$M)

knitr::kable(
  M_scalar_comparison[, c("sex", "method", "estimator_type", "M", "annual_survival")],
  digits = 3,
  caption = "Comparison of scalar empirical natural-mortality estimators"
)
```

```{r}
ggplot(M_scalar_comparison, aes(M, method)) +
  geom_point(size = 2.8) +
  facet_wrap(~ sex) +
  labs(
    title = "Natural mortality depends on the estimator and its inputs",
    subtitle = "Disagreement among empirical methods should be treated as information, not averaged away automatically",
    x = "Estimated M (year⁻¹)", y = NULL
  ) +
  theme_minimal(base_size = 12)
```

### A length-dependent alternative

Gislason et al. (2010) estimate M as a function of body length, growth and asymptotic length. This produces an age/length-dependent mortality pattern once a growth curve is supplied, rather than a single scalar M.

```{r}
M_gislason_curve <- do.call(rbind, lapply(seq_len(nrow(M_inputs)), function(i) {
  d <- M_inputs[i, ]
  out <- M_gislason_at_age(
    age = 1:9,
    Linf_cm = d$Linf,
    K = d$K,
    t0 = d$t0
  )
  out$sex <- d$sex
  out
}))

ggplot(M_gislason_curve, aes(age, M, group = sex, linetype = sex)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  labs(
    title = "Gislason et al. (2010) produces a declining M-at-age pattern",
    subtitle = "The pattern inherits uncertainty from the fitted growth curve and from the empirical mortality relationship",
    x = "Age", y = "M (year⁻¹)", linetype = "Sex"
  ) +
  theme_minimal(base_size = 12)
```

::: {.callout-warning}
## Maximum observed age is not automatically longevity

The `tmax` value used above is deliberately taken from the simulated ageing sample so the calculation is reproducible. In a real stock, the maximum observed age can be truncated by fishing, sampling design, ageing error or limited sample size. Longevity-based estimators should therefore use a defensible biological `tmax`, not simply `max(age)` without review.
:::

### What about PRODBIOM?

PRODBIOM is **not** approximated here by a simple scalar equation. It is a mortality-at-age/length approach based on balancing biomass production and losses under life-history assumptions. Because its implementation is structurally different from the empirical scalar estimators above, reproducing it properly requires a dedicated, validated implementation rather than assigning the label “PRODBIOM” to an arbitrary M vector. It is therefore retained as a candidate future extension, particularly relevant to Mediterranean demersal assessments.

### Methods represented here

| Method | Main inputs | Output | Important caution |
|---|---|---|---|
| Pauly (1980) | Linf, K, mean temperature | scalar M | sensitive to growth and temperature inputs |
| Hoenig (1983) | maximum age | longevity-based mortality relationship | original relationship predicts Z; using it as M requires caution |
| Then et al. (2015), tmax | maximum age | scalar M | depends strongly on defensible longevity |
| Then et al. (2015), growth | Linf, K | scalar M | empirical prediction, not direct estimation |
| Gislason et al. (2010) | L, Linf, K | length-dependent M | inherits growth uncertainty and assumes the empirical size relationship applies |

References used for the implemented equations include Pauly (1980), *Journal du Conseil* 39:175–192, doi:10.1093/icesjms/39.2.175; Gislason et al. (2010), *Fish and Fisheries* 11:149–158, doi:10.1111/j.1467-2979.2009.00350.x; and Then et al. (2015), *ICES Journal of Marine Science* 72:82–92, doi:10.1093/icesjms/fsu136, including the 2018 correction clarifying the length-unit convention for the growth-based equation.

## 10. Use these routines with your own stock

The reusable functions are separated from the simulated dataset in `natural_mortality_methods.R`. A user does not need to reproduce the simulation to use the estimators. With stock-specific biological inputs, the basic pattern is:

```{r eval=FALSE}
source("natural_mortality_methods.R")

# Example only: replace with reviewed stock-specific inputs.
Linf_cm <- 42.5
K <- 0.31
t0 <- -0.4
tmax_years <- 9
mean_temperature_C <- 18

compare_scalar_M(
  Linf_cm = Linf_cm,
  K = K,
  temperature_C = mean_temperature_C,
  tmax_years = tmax_years
)

M_gislason_at_age(
  age = 1:8,
  Linf_cm = Linf_cm,
  K = K,
  t0 = t0
)
```

The important step is **not** replacing the example numbers. It is documenting where each stock-specific input came from, its sex/spatial/temporal scope, and whether the assumptions of each estimator are appropriate. The same principle will progressively be applied to the other modules so that the repository becomes both a worked portfolio and a practical guide for analysts adapting the workflow to their own data.

## 11. Cross-component biological coherence

Life-history components should not be audited independently. Growth, maturity, weight–length, spawning and mortality assumptions need to tell a biologically coherent story.

```{r}
knitr::kable(coherence_summary, digits = 3,
  caption = "Cross-component life-history coherence diagnostics")
```

These are **diagnostics**, not hard biological laws. Their role is to force explicit comparison among inputs before they are embedded in an assessment.

## 12. Assessment-readiness interpretation

The main conclusions from this simulated case are methodological:

- record-level QC must precede fitted biological relationships;
- sampling support and composition should accompany every parameter estimate;
- sex-specific structure can matter for growth, weight and maturity;
- maturity estimates can shift because the observation protocol changes;
- spawning phenology should be checked for temporal and spatial coherence;
- historical or adjacent-stock parameters should not be transferred merely because they are available;
- natural mortality should be compared across defensible estimators and propagated as sensitivity when it cannot be resolved;
- life-history parameters should be reviewed **as a coherent set**, not as independent constants.

::: {.callout-important}
## No assessment model is fitted here

This module stops at the point where the biological evidence can be described as supported, weak, conflicting or sensitive. Selecting and fitting an assessment model remains downstream of this evidence audit.
:::

## Reproducibility

All observations and parameter sources in this module are simulated. The deliberate problems are documented in the data dictionary and generated reproducibly in `06_life_history_consistency.R`. The natural-mortality equations are separated into `natural_mortality_methods.R` so they can be reused with reviewed stock-specific inputs.
